\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap

### 2017-03-19 TeX/paper.tex
scnetwork and coverage probability

 \section*{The Exponential Random Graph Model}
The \textit{exponential random graph model} (ERGM) is a probability model for directed or undirected binary networks. This means neither the weighting nor the temporal change of ties is considered in the model.
In literature, ERGMs are sometimes also referred to as \textit{p-star} or $p^*$ models (see Wassermann and Pattison \cite{Wasserman.1996}). In the following, we will focus on directed networks, however, undirected networks can be introduced very similar.\\[0.3cm] %This chapter is mainly based on Harris \cite{Harris.}, Hunter et al. \cite{Hunter.2008} and Hunter \cite{Hunter.}. \\[0.3cm]
The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$. This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{R}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$. $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected. Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$. Furthermore, define
$$ \mathcal{G}(N) := \left\{ G \in \mathbb{R}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. Therefore, for an already small number of actors the cardinalty of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable. As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods.\\  
With the definition of $\mathcal{G}(N)$ we define 
$$Y: \Omega \to \mathcal{G}(N)~~,~~\omega \mapsto (Y_{ij}(\omega))_{i,j=1,\dots,N}$$
as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N_V)$ we define the ERGM as
\begin{equation}
\mathbb{P}_{\theta}(Y=G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
\label{ERGM}
\end{equation}
%
where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
%
As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
A key role when modeling an ERGM is played by the function $\Gamma(\cdot)$. The decision about which network statistics are incorporated into the model affects the model significantly (see Handcock \cite{Handcock.2003}).

@@

@@

\item include plot
\item Describe plots
\item describe advantage over MPLE (better coverage) and MCMCMLE (faster)
\item can easily be parallelized/ using multiple cores -> increases speed
\item this method is also able to detect degeneracy

@@

\cite{FowlerJeon2008}, \cite{Fowleretal2007}

>>>>

 \section*{The Exponential Random Graph Model}
The \textit{exponential random graph model} (ERGM) is a probability model for directed or undirected binary networks. This means neither the weighting nor the temporal change of ties is considered in the model.
In literature, ERGMs are sometimes also referred to as \textit{p-star} or $p^*$ models (see Wassermann and Pattison \cite{Wasserman.1996}, Robins et al. \cite{robins.pattison.2007}). In the following, we will focus on directed networks, however, undirected networks can be introduced very similar.\\[0.3cm] %This chapter is mainly based on Harris \cite{Harris.}, Hunter et al. \cite{Hunter.2008} and Hunter \cite{Hunter.}. \\[0.3cm]
The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$. This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{R}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$. $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected. Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$. Furthermore, define
$$ \mathcal{G}(N) := \left\{ G \in \mathbb{R}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. Therefore, for an already small number of actors the cardinalty of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable. As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods.\\  
With the definition of $\mathcal{G}(N)$ we define 
$$Y: \Omega \to \mathcal{G}(N)~~,~~\omega \mapsto (Y_{ij}(\omega))_{i,j=1,\dots,N}$$
as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N_V)$ we define the ERGM as
\begin{equation}
\mathbb{P}_{\theta}(Y=G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
\label{ERGM}
\end{equation}
%
where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
%
As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
A key role when modeling an ERGM is played by the function $\Gamma(\cdot)$. The decision about which network statistics are incorporated into the model affects the model significantly (see Handcock \cite{Handcock.2003}).

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.4]{Coverage_faux_mesa_high}
\caption{The Coverage Probability results of the Faux Mesa High network for bootstrapped MPLE, MCMCMLE and logistic regression }
\label{rmse}
\end{center}
\end{figure}

@@

\item include magnolia plot
\item Describe plots
\item describe advantage over MPLE (better coverage) and MCMCMLE (faster)
\item can easily be parallelized/ using multiple cores -> increases speed
\item this method is also able to detect degeneracy

@@

In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMCMLE we apply both approaches to the U.S. Supreme Court Citation Network developed by Fowler and Jeon \cite{FowlerJeon2008} and Fowler et al. \cite{Fowleretal2007}. The citation network we use consists of 9223 majority opinions written by the U.S. Supreme Court from 1954 to 2002. In this undirected network an edge occurs between two verticies (i.e. between two majority opinions) if one majority opinion cites the other. The endogenous statistics we include into this model are the number of edges, the edge-wise 0-shared partner (esp(0)) statistic and the dyad-wise 0-shared partner statistic (dsp(0)) (Snijders et al. \cite{SnijdersTomA.B..2006}, Hunter and Handcock \cite{Hunter.2006}). The esp(k) statistic counts the number of pairs which are directly linked and share exactly k partners. The dsp(k) statistic is quite similar to esp(k), but does not require the two vertices to be directly linked. ({\it add explanation for adding esp(0) and dsp(0)}). 
The two nodal attributes we consider for this network are the year the majority opinion was drafted as well as an attribute that indicates whether a case appears on the Oxford list of salient cases. Beside these two nodal attributes we furthermore include the difference between the years of publication as an edge attribute. Since we do expect a non-linear effect fot this edge attribute, we include this variable as a second-degree polynomial.
